Quasi-Monte Carlo Methods: What, Why, and How?

### 2025-01-20 HickernellMCQMC2024_QMCTutorial.tex
M2 Mac updates
Updates from Overleaf

It is commonplace in qMC constructions that equidistribution is required not just in the (possibly high dimensional) $d$-dimensional space, but also for all or specific low-dimensional projections; see Section \ref{sec:coordwts}. MatBuilder \cite{paulin2022} is a software tool developed precisely for this task; to generate digital sequence LD nodes when low-dimensional uniformity is critical. Uniformity constraints related to digital sequence stratification across boxes of the form \eqref{eq:element_box} are expressed as integer linear programming (ILP) problems and subsequently solved to optimize an objective function relating to the ILP reformulation. The solutions then determine the next columns in the generator matrices for the digital sequence.

As discussed above, the Halton sequence has the benefit of not having a preferred number of points for its construction. It does however possess the drawback of exhibiting poor uniformity in the lower dimensional projections when $d$ is moderate to high due to intrinsic number theoretic correlation between the coordinates \cite{kirklem24}. One popular strategy to circumvent this poor behavior, dating back to Faure \cite{fau78} and Braaten and Weller \cite{brawell79}, is to randomly or deterministically permute the digits of the base $b_j$ expansions $j = 1,2\ldots,d$ to break the observed correlation between the components. To do so, we first define the generalized van der Corput sequence by obtaining a sequence $\Sigma = (\sigma_r)_{r\geq0}$ of permutations of $\{0,\ldots,b-1\}$ where the $i^{th}$ term of the one-dimensional sequence is defined as 
\begin{multline}\label{eq:gen_vdc}
    \phi_b^{\Sigma}(i_0 + i_1b + i_2 b^2 + \cdots ) := \sigma_0(i_0) b^{-1} + \sigma_1(i_1) b^{-2} + \sigma_2(i_2) b^{-3} + \cdots \in [0,1)
	\\
	 \text{where } i_0, \sigma_0(i_0), i_1, \sigma_1(i_1), \ldots \in \{0,\ldots b-1\}.
\end{multline}
The generalized Halton sequence is subsequently defined by choosing $d$ sequences of permutations $\Sigma_j = (\sigma_{j,r})_{r\geq0}, j = 1, 2,\ldots,d$ and setting
$$\bsx_i^{\text{G-Hal}} = \left( \phi^{\Sigma_1}_{b_1}(i), \ldots, \phi^{\Sigma_d}_{b_d}(i)\right), \quad i= 0,1,\ldots,$$
There exist many attempts to optimize the permutations defining the generalized sequence, most of which involve so-called linear digit scramblings. We point the reader towards \cite{} and the excellent survey of the work in this direction (until 2008) by Faure and Lemieux \cite{}.

So far, we have discussed the attempts of researchers to optimize specific classes of LD constructions; lattices and digital sequences, and Halton sequences.  There also exist several efforts, some very recently, to generate novel LD constructions which don't follow any number theoretic or any other underlying structure, but none-the-less, possess a high level of uniformity. These methods are often global optimization methods, which is often challenging due to the non-convex nature of the FOM as a function of the nodes. Traditional global optimization methods, such as gradient descent or its variants, struggle in this setting becoming trapped in suboptimal local minima and failing to identify the global optimum configuration of nodes. One of the first works to globally optimize LD nodes, albeit a heuristic approach, was by Winker and Fang \cite{WinFan97b} employing the threshold accepting optimization algorithm to directly optimize the uniformity of the LD nodes with respect to the discrepancy FOM. More recently, Rusch, Kirk, Bronstein, Lemieux and Rus \cite{ruschkirk24} developed the first machine-learning assisted optimization tool for LD node optimization in the Message-Passing Monte Carlo (MPMC) method. MPMC implements tools from geometric deep learning imposing a computational graph onto an initialized IID node set to transform the nodes to LD node set via a learned mapping, again guided by the minimization of a discrepancy-based FOM. We also highlight the method proposed by Hinrichs and Oettershagen \cite{hinoet16}, which generates optimal LD nodes for a specific class of functions in two-dimensions. To address the non-convexity challenge, the authors decompose the global optimization of a specific FOM into a large number of smaller, convex sub-problems which are solved individually. There are several other attempts to construct optimal LD nodes, many of which are well summarized in the thesis of Cl\'{e}ment \cite{}. We highlight the subset selection method proposed in \cite{cle22} to choose from an $n$-element initialization set, the $k<n$ nodes which minimize a discrepancy FOM. A heuristic approach to this problem was later shown in \cite{cle24}. Furthermore, a method to generate optimal star-discrepancy node sets for fixed $n$ and $d$ based on a non-linear programming approach was suggested in \cite{}.

>>>>

It is commonplace in qMC constructions that equidistribution is required not just in the (possibly high dimensional) $d$-dimensional space, but also for all or specific low-dimensional projections. MatBuilder \cite{paulin2022} is a software tool developed precisely for this task; to generate digital sequence LD nodes when low dimensional uniformity is critical. Uniformity constraints related to digital sequence stratification across boxes of the form \eqref{eq:element_box} are expressed as integer linear programming (ILP) problems and subsequently solved to optimize an objective function relating to the ILP reformulation ensuring good quality low dimensional projections. The solutions determine then build the columns of the generator matrices in a greedy manner for the LD digital sequence.

As discussed in the previous section, the Halton sequence has the benefit of not having a preferred number of points for its construction. It does however possess the drawback of exhibiting poor uniformity in the lower dimensional projections when $d$ is moderate to high due to intrinsic number theoretic correlation between the coordinates \cite{kirklem24}. One popular strategy to circumvent this poor behavior, dating back to Faure \cite{fau78} and Braaten and Weller \cite{brawell79}, is to randomly or deterministically permute the digits of the base $b_j$ expansions $j = 1,2\ldots,d$ to disrupt correlations between the components. To do so, we first define the generalized van der Corput sequence by obtaining a sequence $\Sigma = (\sigma_r)_{r\geq0}$ of permutations of $\{0,\ldots,b-1\}$ where the $i^{th}$ term of the one-dimensional sequence is defined as 
\begin{multline}\label{eq:gen_vdc}
    \phi_b^{\Sigma}(i_0 + i_1b + i_2 b^2 + \cdots ) := \sigma_0(i_0) b^{-1} + \sigma_1(i_1) b^{-2} + \sigma_2(i_2) b^{-3} + \cdots \in [0,1)
	\\
	 \text{where } i_0, \sigma_0(i_0), i_1, \sigma_1(i_1), \ldots \in \{0,\ldots b-1\}.
\end{multline}
The generalized Halton sequence is subsequently defined by choosing $d$ sequences of permutations $\Sigma_j = (\sigma_{j,r})_{r\geq0}, j = 1, 2,\ldots,d$ and setting
$$\bsx_i^{\text{G-Hal}} = \left( \phi^{\Sigma_1}_{b_1}(i), \ldots, \phi^{\Sigma_d}_{b_d}(i)\right), \quad i= 0,1,\ldots,$$
There exist many proposed algorithms to optimize the permutations defining the generalized sequence, most of which involve so-called linear digit scramblings and usually involve directly optimizing a discrepancy FOM. We point the reader towards \cite{MasChiWar05,tuffin98,Atan04,vancools06} and the excellent survey of the work in this direction (until 2008) by Faure and Lemieux \cite{faulem09}.

So far, we have discussed the attempts of researchers to optimize specific classes of LD constructions; lattices and digital sequences, and Halton sequences.  There also exist several efforts, some very recently, to generate novel LD constructions which don't follow any number theoretic or any other underlying structure, but none-the-less, possess a high level of uniformity. These methods are often global optimization methods, which is often challenging due to the non-convex nature of the FOM as a function of the nodes. Traditional global optimization methods, such as gradient descent or its variants, struggle in this setting becoming trapped in suboptimal local minima and failing to identify the global optimum configuration of nodes. One of the first works to globally optimize LD nodes, albeit a heuristic approach, was by Winker and Fang \cite{WinFan97b} employing the threshold accepting optimization algorithm to directly optimize the uniformity of the LD nodes with respect to a discrepancy FOM. More recently, Rusch, Kirk, Bronstein, Lemieux and Rus \cite{ruschkirk24} developed the first machine-learning assisted optimization tool for LD node optimization in the Message-Passing Monte Carlo (MPMC) method. MPMC implements tools from geometric deep learning imposing a computational graph onto an initialized IID node set to transform the nodes to LD node set via a learned mapping, again guided by the minimization of a discrepancy-based FOM. We also highlight the method proposed by Hinrichs and Oettershagen \cite{hinoet16}, which generates optimal LD nodes for a specific class of functions in two-dimensions. To address the non-convexity challenge, the authors decompose the global optimization of a specific FOM into a large number of smaller, convex sub-problems which are solved individually. There are several other recent attempts to construct optimal LD nodes, many of which are well summarized in the thesis of Cl\'{e}ment \cite{clethesis24}. We highlight the subset selection method proposed in \cite{cle22} to choose from an $n$-element initialization set, the $k<n$ nodes which minimize a discrepancy FOM. A heuristic approach to this problem was later shown in \cite{cle24}. Furthermore, a method to generate optimal star-discrepancy node sets for fixed $n$ and $d$ based on a non-linear programming approach was suggested in \cite{cle24}, however is computationally restricted to very small numbers of points in dimensions two and three only.
